# Interactive `ggtree` {#sec-interactive-ggtree}

```{r}
library("ape")
library("ggtree")
source("setup.R")
```

## Introduction

Tree figures are often used in two different modes.

In the first mode, you prepare a static figure for a manuscript or a report. The goal is to be stable and reproducible. In the second mode, you explore: you hover to read values, identify elements, and verify that layers match the intended nodes or clades.

`r Biocpkg("ggtree")` supports an interactive mode designed for exploration. The design principle is simple: you switch interactivity on and off, while keeping the same plotting code and the same `ggtree` objects. When interactivity is enabled, printing a plot returns an HTML widget; when it is disabled, the same plot prints as a static figure.

## Turning interactivity on and off

Start with a normal `ggtree` plot.

```{r}
set.seed(123)
tr <- rtree(20)

p <- ggtree(tr)
p
```

Enable interactive mode and print the same object again.

```{r}
ggtree_set_interactive()
p
```

At this stage, the output may still look “plain”. This is expected: interactivity requires you to tell `ggtree` what should be shown on hover and how graphical elements should be identified.

When you are done exploring and want static output again, disable interactive mode.

```{r}
ggtree_unset_interactive()
p
```

## The two key aesthetics: `tooltip` and `data_id`

Interactive behavior is controlled by two aesthetics:

- `tooltip` supplies the content shown when hovering over an element.
- `data_id` supplies a unique identifier for elements so that the widget can map hover/click events back to specific branches, nodes, or layers.

The following example uses branch length as the tooltip and uses node IDs as element identifiers.

```{r}
ggtree_set_interactive()

p2 <- ggtree(
  tr,
  mapping = aes(
    tooltip = round(branch.length, 2),
    data_id = node
  )
)

p2
```

With these two mappings, hovering over branches should display the tooltip, and the hovered element will be visually emphasized.

## Interactive layers

You can add interactive behavior to additional layers in the same way: by supplying `tooltip` and `data_id` in the layer mapping. This is useful when you highlight clades, place symbols, or add annotations and want them to participate in the interactive view.

```{r}
dt <- data.frame(id = c(36, 38), type = c("A", "B"))

p3 <- p2 +
  geom_hilight(
    data = dt,
    mapping = aes(
      node = id,
      fill = type,
      tooltip = paste0("clade of node ", id),
      data_id = type
    ),
    to.bottom = TRUE
  )

p3
```

In interactive mode, hover emphasis is applied on top of the original colors. If a highlighted region appears to mismatch the legend while hovering, it is usually the hover overlay. The underlying colors remain unchanged.

## Returning to static output

Interactive plots are intended for HTML output. For PDF export and static raster/vector formats, disable interactive mode and print the same plot.

```{r}
ggtree_unset_interactive()
p3
```

This workflow allows you to use interactivity during exploration while keeping a single plotting script for both interactive inspection and final figure export.

## Practical notes

- Assign stable and unique `data_id` values within each layer.
- Keep tooltips short. Use them for values, labels, and short summaries.
- If a plot becomes slow, remove unnecessary interactive mappings from dense layers.
